---
title: "Using the Example Omics Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Example Omics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette runs `prepare_metabolites()` and `prepare_proteins()` end to end
using the example files bundled with the package in `inst/extdata`.

> **The bundled files are synthetic.** Every value was drawn from a random
> number generator by `data-raw/make-example-data.R`. They contain no real
> measurements and no human subject data. Their column names and layout imitate
> a cleaned Progenesis QI peak table and a wide precursor table so that the
> examples look like real input.

```{r load-package}
library(Multiomics.Preprocessing)
```

## Locate and read the example files

```{r read-extdata}
extdata_dir <- system.file("extdata", package = "Multiomics.Preprocessing")

# Shared sample metadata used by both workflows.
samples <- read.csv(file.path(extdata_dir, "samples.csv"), check.names = FALSE)

# Four metabolomics peak tables (two ionisation modes x two runs).
neg_m1 <- read.csv(file.path(extdata_dir, "Metabolites_NegM1.csv"), check.names = FALSE)
neg_m2 <- read.csv(file.path(extdata_dir, "Metabolites_NegM2.csv"), check.names = FALSE)
pos_m1 <- read.csv(file.path(extdata_dir, "Metabolites_PosM1.csv"), check.names = FALSE)
pos_m2 <- read.csv(file.path(extdata_dir, "Metabolites_PosM2.csv"), check.names = FALSE)

# Three proteomics fractions, already pivoted to one column per sample.
fraction_1 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_1.csv"), check.names = FALSE)
fraction_2 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_2.csv"), check.names = FALSE)
fraction_3 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_3.csv"), check.names = FALSE)

str(samples)
```

Sample identifiers live in `samples$Sample_ID`, and the abundance columns in
each omics table are named after those identifiers. Both functions detect the
metadata column holding the sample IDs automatically; if more than one column
qualifies, they stop and ask you to name it with `sample_id_column`.

## Selecting the abundance columns

The peak and precursor tables begin with annotation columns and then carry one
column per sample. Select the abundance columns **by name** rather than by a
numeric range: a range silently rots if the file layout changes, and picking up
an annotation column by mistake is the most common way to get an error here.

```{r pick-columns}
abundance_columns <- intersect(samples$Sample_ID, colnames(neg_m1))
length(abundance_columns)
setdiff(colnames(neg_m1), abundance_columns)   # the annotation columns
```

## Prepare the metabolomics data

Mixed imputation classifies each peak by whether its *detection* pattern is
explained by `Sample_group` and `Sex`, then imputes accordingly: QRILC for peaks
that look left-censored, random forest for the rest. Imputation happens
separately within each metabolomics batch.

```{r prepare-metabolites}
metabolite_results <- prepare_metabolites(
  input_files = list(neg_m1, neg_m2, pos_m1, pos_m2),
  dataset_file_names = c("NegM1", "NegM2", "PosM1", "PosM2"),
  peak_abundance_columns = abundance_columns,
  peak_id_column = "Compound",
  sample_meta_data = samples,
  zero_handling = "na",
  impute_method = "mixed",
  mnar_variables = c("Sample_group", "Sex"),
  mnar_significance_threshold = 0.01,
  imputation_seeds = 1234,
  impute_behaviour = "per batch",
  missing_feature_proportion_threshold = c(0.3, 0.75),
  normalisation_method = "vsn",
  batch_correction = TRUE,
  batch_column = "Metabolomic_batch",
  batch_correction_variable_columns = c("Sample_group", "Sex"),
  metabolite_aggregator_method = MsCoreUtils::robustSummary,
  metabolite_aggregator_column = "Compound_ID",
  verbose = FALSE
)
```

```{r inspect-metabolites}
names(metabolite_results)

# The processed metabolite matrix.
dim(metabolite_results$metabolites)
round(metabolite_results$metabolites[1:5, 1:6], 2)

# Missingness classification for the combined peak table.
head(metabolite_results$mnar_results[["raw"]])
table(metabolite_results$mnar_results[["raw"]]$MNAR)
```

`MNAR` marks peaks whose detection is associated with the covariates you
supplied. That is a heuristic for routing peaks to left-censored imputation, not
a proof of the formal missing-not-at-random mechanism, which depends on values
that were never measured.

## Prepare the proteomics data

The three fractions deliberately carry different ancillary annotation columns
(`Modified.Sequence`, `Proteotypic`, `Precursor.Charge`). They are combined by
taking the union, so a fraction that lacks a column gets `NA` for it.

```{r prepare-proteins}
protein_results <- prepare_proteins(
  input_files = list(fraction_1, fraction_2, fraction_3),
  fraction_file_names = c("Fraction_1", "Fraction_2", "Fraction_3"),
  precursor_abundance_columns = abundance_columns,
  precursor_id_column = "Precursor.Id",
  sample_meta_data = samples,
  zero_handling = "leave",
  impute_method = "mixed",
  mnar_variables = c("Sample_group", "Sex"),
  imputation_seeds = 5678,
  impute_behaviour = "global",
  missing_feature_proportion_threshold = c(0.3, 0.75),
  batch_specific_features_threshold = 0.9,
  normalisation_method = "vsn",
  batch_correction = TRUE,
  batch_column = "Proteomic_batch",
  batch_correction_variable_columns = c("Sample_group", "Sex"),
  protein_aggregator_method = MsCoreUtils::robustSummary,
  protein_aggregator_column = "Protein.Group",
  verbose = FALSE
)
```

```{r inspect-proteins}
names(protein_results)

dim(protein_results$proteins)
round(protein_results$proteins[1:5, 1:6], 2)

# One classification table per fraction.
names(protein_results$mnar_results)
```

## Notes

- Long-format search engine output (a DIA-NN `report.tsv`, for example) must be
  pivoted to one column per sample before it is passed to `prepare_proteins()`.
- `impute_method = "mixed"` needs the `missForest` and `imputeLCMD` packages.
  They are optional dependencies; the function checks for them before doing any
  work and tells you how to install them.
- Set `plot_qc_charts = TRUE` for density plots and before/after batch-correction
  PCA. These are captured with `grDevices::recordPlot()` and are intended for
  interactive sessions.
- Set `imputation_seeds` for reproducible imputation. The caller's random number
  generator is restored when the function returns, so seeding here does not
  affect the rest of your script.
- If you only want the abundance matrix, set `return_mnar_results = FALSE`.
