---
title: "Using the Example Omics Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Example Omics Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette shows how to run `prepare_metabolites()` and `prepare_proteins()`
using the example files that are bundled with the package in `inst/extdata`.

```{r load-package, eval = FALSE}
# Load the package.
library(Multiomics.Preprocessing)
```

## Locate and read the example files

```{r read-extdata, eval = FALSE}
# Find the installed extdata folder so the code works after the package is installed.
extdata_dir <- system.file("extdata", package = "Multiomics.Preprocessing")

# Read the shared sample metadata used by both workflows.
samples <- read.csv(file.path(extdata_dir, "samples.csv"), check.names = FALSE)

# Read the four metabolomics example files.
neg_m1 <- read.csv(file.path(extdata_dir, "Metabolites_NegM1.csv"), check.names = FALSE)
neg_m2 <- read.csv(file.path(extdata_dir, "Metabolites_NegM2.csv"), check.names = FALSE)
pos_m1 <- read.csv(file.path(extdata_dir, "Metabolites_PosM1.csv"), check.names = FALSE)
pos_m2 <- read.csv(file.path(extdata_dir, "Metabolites_PosM2.csv"), check.names = FALSE)

# Read the three proteomics fraction files.
fraction_1 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_1.csv"), check.names = FALSE)
fraction_2 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_2.csv"), check.names = FALSE)
fraction_3 <- read.csv(file.path(extdata_dir, "Proteomics_fraction_3.csv"), check.names = FALSE)
```

The `samples` table contains the sample-level metadata used during MNAR
classification, optional batch-aware processing, and optional batch correction.
The functions automatically detect which metadata column contains the sample IDs
by matching the values in `sample_meta_data` to the abundance column names in
the omics tables.

## Optional quick trial

If you want a faster test run before processing the full example files, you can
subset the feature tables first.

```{r optional-trim, eval = FALSE}
# Keep only the first 500 features from each file for a quick trial run.
neg_m1_small <- neg_m1[1:500, , drop = FALSE]
neg_m2_small <- neg_m2[1:500, , drop = FALSE]
pos_m1_small <- pos_m1[1:500, , drop = FALSE]
pos_m2_small <- pos_m2[1:500, , drop = FALSE]

fraction_1_small <- fraction_1[1:500, , drop = FALSE]
fraction_2_small <- fraction_2[1:500, , drop = FALSE]
fraction_3_small <- fraction_3[1:500, , drop = FALSE]
```

## Prepare the metabolomics data

```{r prepare-metabolites, eval = FALSE}
# Run the metabolomics preprocessing workflow on the four example datasets.
# Here we use mixed imputation, classify MNAR features using Sample_group and Sex,
# and impute within each metabolomics batch.
metabolite_results <- prepare_metabolites(
  input_files = list(neg_m1, neg_m2, pos_m1, pos_m2),
  dataset_file_names = c("NegM1", "NegM2", "PosM1", "PosM2"),
  peak_abundance_columns = 5:ncol(neg_m1),
  peak_id_column = "Compound",
  sample_meta_data = samples,
  zero_handling = "na",
  impute_method = "mixed",
  mnar_variables = c("Sample_group", "Sex"),
  mnar_significance_threshold = 0.01,
  return_mnar_results = TRUE,
  impute_behaviour = "per batch",
  missing_feature_proportion_threshold = c(0.3, 0.75),
  batch_specific_features_threshold = NULL,
  normalisation_method = "vsn",
  batch_correction = FALSE,
  batch_column = "Metabolomic_batch",
  batch_correction_variable_columns = c("Sample_group", "Sex"),
  metabolite_aggregator_method = MsCoreUtils::robustSummary,
  metabolite_aggregator_column = "Compound ID",
  plot_qc_charts = FALSE,
  verbose = TRUE
)
```

```{r inspect-metabolites, eval = FALSE}
# Because return_mnar_results = TRUE, the result is a list.
names(metabolite_results)

# The processed metabolite matrix is stored in the metabolites element.
dim(metabolite_results$metabolites)
head(metabolite_results$metabolites[, 1:6])

# MNAR classification tables are returned for the input datasets.
names(metabolite_results$mnar_results)
head(metabolite_results$mnar_results[[1]])
```

## Prepare the proteomics data

```{r prepare-proteins, eval = FALSE}
# Run the proteomics preprocessing workflow on the three fraction files.
# Here we again use mixed imputation, but perform imputation globally across
# all samples before aggregating precursor features to proteins.
protein_results <- prepare_proteins(
  input_files = list(fraction_1, fraction_2, fraction_3),
  fraction_file_names = c("Fraction_1", "Fraction_2", "Fraction_3"),
  precursor_abundance_columns = 5:ncol(fraction_1),
  precursor_id_column = "Precursor.Id",
  sample_meta_data = samples,
  zero_handling = "na",
  impute_method = "mixed",
  mnar_variables = c("Sample_group", "Sex"),
  mnar_significance_threshold = 0.01,
  return_mnar_results = TRUE,
  impute_behaviour = "global",
  missing_feature_proportion_threshold = c(0.3, 0.75),
  normalisation_method = "vsn",
  batch_correction = FALSE,
  batch_column = "Proteomic_batch",
  batch_correction_variable_columns = c("Sample_group", "Sex"),
  protein_aggregator_method = MsCoreUtils::robustSummary,
  protein_aggregator_column = "Protein.Group",
  plot_qc_charts = FALSE,
  verbose = TRUE
)
```

```{r inspect-proteins, eval = FALSE}
# Because return_mnar_results = TRUE, the result is a list.
names(protein_results)

# The processed protein matrix is stored in the proteins element.
dim(protein_results$proteins)
head(protein_results$proteins[, 1:6])

# MNAR classification tables are returned for the input fractions.
names(protein_results$mnar_results)
head(protein_results$mnar_results[[1]])
```

## Notes

- Use `impute_behaviour = "per batch"` when you want missing values to be
  imputed separately within each batch.
- Use `batch_correction = TRUE` if you want ComBat correction after
  normalisation.
- Set `plot_qc_charts = TRUE` if you want density and PCA QC plots returned in
  the output object.
- If you only want the processed abundance matrix, set
  `return_mnar_results = FALSE`.
